---
title: "Box-plots"
author: "Chris Hanretty"
date: "November 2020"
output:
  xaringan::moon_reader:
    css: ["default", "rhul.css", "rhul-fonts.css","https://cdnjs.cloudflare.com/ajax/libs/animate.css/3.7.0/animate.min.css"]
    nature:
          highlightLines: true
---


```{r knitropts, echo = FALSE, message = FALSE}

knitr::opts_chunk$set(dpi=300,fig.width=4.5, fig.height = 4, warning = TRUE, message = TRUE, out.width = "100%")

```

```{r loadlibs, echo = FALSE, message = FALSE}
library(tidyverse)
library(flipbookr)
```

# About box-plots

 - A **lot** more common in academic data-viz than elsewhere
 - These show the distribution of a continuous outcome according to another (categorical) variable
 - You do need to learn to read box-plots, and what the boxes and
   mid-points represent
 - Alternatives to box-plots include **ridge-plots**, **violin plots**, and **beeswarm plots**

```{r gendata, echo = FALSE}
set.seed(2033)
dat <- data.frame(fac = "a",
                  y = rt(250, 5, 0))

```

```{r egplot, echo = FALSE, eval = FALSE}
ggplot(dat) +
    aes(x = fac) +
    aes(y = y) +
    geom_boxplot(width = 1/3) +
    scale_x_discrete("Some categorical variable") +
    scale_y_continuous("Some continuous outcome") +
    annotate("text",
             x = 1.4,
             y = median(dat$y),
             label = "Median") + 
    annotate("text",
             x = 1.4,
             y = quantile(dat$y, 0.75),
             label = "75th percentile") +
    annotate("text",
             x = 1.4,
             y = quantile(dat$y, 0.25),
             label = "25th percentile") + 
    annotate("text",
             x = 1.4,
             y = median(dat$y) + 1.5 * IQR(dat$y),
             label = "Median + 1.5 times\ninterquartile range") +
    annotate("text",
             x = 1.4,
             y = median(dat$y) - 1.5 * IQR(dat$y),
             label = "Median - 1.5 times\ninterquartile range") +
    annotate("text",
             x = 1.4,
             y = max(dat$y),
             label = "Outlier")
```
 
---


`r chunk_reveal(chunk_name = "egplot", break_type = "auto", title = "# What everything means")`

---

# A realistic example

 - Let's show the spread of changes in the Conservative vote (2017 to 2019) according to region
 - **Change in the Conservative vote** is our continuous outcome
 - **Region** is our categorical predictor
 - We use the same BES results file as we used for scatterplots
 
```{r besin}
library(tidyverse)
library(rio)
bes <- rio::import("https://www.britishelectionstudy.com/wp-content/uploads/2020/02/BES-2019-General-Election-results-file-v1.0.xlsx")

bes <- bes %>%
    mutate(ConChg = Con19 - Con17) %>%
    select(ConstituencyName, Region, ConChg)
           
```

```{r conchgplot, echo = FALSE, eval = FALSE}
ggplot(bes) +
    aes(x = Region) +
    aes(y = ConChg) +
    xlab("Region") +
    ylab("Change in Conservative vote\n2017 to 2019") +
    geom_boxplot() +
    labs(title = "Change in the Conservative vote\nby region") +
    coord_flip() +
    theme_bw()
```

---

`r chunk_reveal(chunk_name = "conchgplot", break_type = "auto")`

---

# Improving the plot

At the moment, regions are listed in reverse alphabetical order, because... reasons.

Let's change that using the `fct_reorder()` function.

```{r fctreorder}
bes <- bes %>%
    mutate(Region = fct_reorder(Region,
                                ConChg,
                                .fun = median,
                                na.rm = TRUE))
```

```{r conchgplot2, echo = FALSE, eval = FALSE}
ggplot(bes) +
    aes(x = Region) +
    aes(y = ConChg) +
    xlab("Region") +
    ylab("Change in Conservative vote\n2017 to 2019") +
    geom_boxplot() +
    labs(title = "Change in the Conservative vote\nby region") +
    coord_flip() +
    theme_bw()
```

---

`r chunk_reveal(chunk_name = "conchgplot2", break_type = "auto")`

---

# Alternative visualizations

Let's play around with some different visualizations which might look more appealling.

To do so, we'll need to load an additional package.

This package will not be installed by default, so you'll need to run
`install.packages("ggbeeswarm")` before running this next batch of
code.

```{r loadvizlibs}
library(ggridges)
library(ggbeeswarm)
```


```{r violin, echo = FALSE, eval = FALSE}
ggplot(bes) +
    aes(x = Region) +
    aes(y = ConChg) +
    xlab("Region") +
    ylab("Change in Conservative vote\n2017 to 2019") +
    geom_violin() +
    labs(title = "Change in the Conservative vote\nby region") +
    coord_flip() +
    theme_bw()
```


```{r ridge, echo = FALSE, eval = FALSE}
ggplot(bes) +
    aes(x = Region) +
    aes(y = ConChg) +
    xlab("Region") +
    ylab("Change in Conservative vote\n2017 to 2019") +
    geom_density_ridges() +
    labs(title = "Change in the Conservative vote\nby region") +
    coord_flip() +
    theme_bw()
```

```{r swarm, echo = FALSE, eval = FALSE}
ggplot(bes) +
    aes(x = Region) +
    aes(y = ConChg) +
    xlab("Region") +
    ylab("Change in Conservative vote\n2017 to 2019") +
    geom_beeswarm(alpha = 0.5) +
    labs(title = "Change in the Conservative vote\nby region") +
    coord_flip() +
    theme_bw()
```

```{r swarmplus, echo = FALSE, eval = FALSE}
ggplot(bes) +
    aes(x = Region) +
    aes(y = ConChg) +
    xlab("Region") +
    ylab("Change in Conservative vote\n2017 to 2019") +
    geom_boxplot(colour = "red") + 
    geom_beeswarm(alpha = 0.5) +
    labs(title = "Change in the Conservative vote\nby region") +
    coord_flip() +
    theme_bw()
```

---

`r chunk_reveal(chunk_name = "violin", break_type = "auto", title = "# Violin plot")`

---


`r chunk_reveal(chunk_name = "swarm", break_type = "auto", title = "# Beeswarm plot")`

---

`r chunk_reveal(chunk_name = "swarmplus", break_type = "auto", title = "# Beeswarm plot (2)")`
