---
title: "Scatterplots"
author: "Chris Hanretty"
date: "November 2020"
output:
  xaringan::moon_reader:
    css: ["default", "rhul.css", "rhul-fonts.css","https://cdnjs.cloudflare.com/ajax/libs/animate.css/3.7.0/animate.min.css"]
    nature:
          highlightLines: true
---


```{r knitropts, echo = FALSE, message = FALSE}

knitr::opts_chunk$set(dpi=300,fig.width=4.5, fig.height = 4, warning = TRUE, message = TRUE, out.width = "100%")

```

# About scatterplots

 - One of the most common data visualizations in the social sciences
 - Plot one continuous variable against another continuous variable
 - By convention, we put the thing we're explaining on the vertical axis
 - The thing we use to explain goes on the horizontal axis
 - Everything else is fair game
 
---

# About this session

 - This session will show you how to construct a scatterplot using `ggplot`
 - It assumes some prior knowledge of how R works
 - It is not a complete guide to ggplot
 - Useful resources include:
    * Chapter 3 of ["R for Data Science"](https://r4ds.had.co.nz/data-visualisation.html)
	* The [ggplot2 manual](https://ggplot2.tidyverse.org/)
	* Kieran Healy's [Data Visualization](https://socviz.co/)

---

# Load the libraries

```{r loadlibs}
library(tidyverse)
```

The `tidyverse` is a container package which includes `ggplot2`.

```{r loadlibs2, echo = FALSE}
library(flipbookr)
```

---

# Get the data in

```{r datin}
dat <- read.csv("gdp_dem_data.csv")
```

Download from `https://raw.githubusercontent.com/chrishanretty/dataviz/main/01_democratization/gdp_dem_data.csv`.

---

# Look at the data


```{r datlook}
head(dat, 5)
```

```{r figcode, eval = FALSE, echo = FALSE}
ggplot(dat) +
    aes(x = gdp_pc_1992) +
    aes(y = v2x_libdem) +
    geom_point() +
    xlab("GDP per capita in 1992, US$") +
    ylab("V-Dem measure of liberal democracy") +
    scale_x_log10(breaks = c(400, 1000,
                             4000, 10000,
                             40000)) +
    labs(title = "Countries richer in 1992 have\nhigher levels of democracy now") +
    labs(caption = "Source: V-Dem, Penn World Tables") + 
    theme_bw()
```

---

`r chunk_reveal(chunk_name = "figcode", break_type = "auto")`

---

# All the code

```{r figcode, eval = FALSE}
```

---

# How I'd usually write it

```{r figcode2, eval = FALSE}
ggplot(dat, aes(x = gdp_pc_1992, y = v2x_libdem)) +
    geom_point() +
    scale_x_log10("GDP per capita in 1992, US$",
                  breaks = c(400, 1000,
                             4000, 10000,
                             40000)) +
    scale_y_continuous("V-Dem measure of liberal democracy")
    labs(title = "Countries richer in 1992 have\nhigher levels of democracy now", 
         caption = "Source: V-Dem, Penn World Tables") + 
    theme_bw()
```

---

# Further options

1. Adding a trend line
2. Mapping an additional aesthetic
3. Small multiples

---

# Adding a trend line

```{r trendline, eval = FALSE}
ggplot(dat) +
    aes(x = gdp_pc_1992) +
    aes(y = v2x_libdem) +
    geom_point() +
    xlab("GDP per capita in 1992, US$") +
    ylab("V-Dem measure of liberal democracy") +
    scale_x_log10(breaks = c(400, 1000,
                             4000, 10000,
                             40000)) +
    labs(title = "Countries richer in 1992 have\nhigher levels of democracy now") +
    labs(caption = "Source: V-Dem, Penn World Tables") + #BREAK
    geom_smooth(method = "lm") + 
    theme_bw()
```

---

`r chunk_reveal(chunk_name = "trendline", break_type = "auto")`


---

# Mapping an additional aesthetic (colour)


```{r addaes, eval = FALSE}
ggplot(dat) +
    aes(x = gdp_pc_1992) +
    aes(y = v2x_libdem) + #BREAK
    aes(colour = e_regiongeo) + #BREAK2
    geom_point() + #BREAK3
    xlab("GDP per capita in 1992, US$") +
    ylab("V-Dem measure of liberal democracy") +
    scale_x_log10(breaks = c(400, 1000,
                             4000, 10000,
                             40000)) +
    labs(title = "Countries richer in 1992 have\nhigher levels of democracy now") +
    labs(caption = "Source: V-Dem, Penn World Tables") + #BREAK4
    scale_colour_discrete("Region") + 
    theme_bw() + #BREAK5
    theme(legend.position = "bottom") 
```

---

`r chunk_reveal(chunk_name = "addaes", break_type = "auto")`

---

# Mapping an additional aesthetic (size)


```{r addaes2, echo = FALSE, eval = FALSE}
ggplot(dat) +
    aes(x = gdp_pc_1992) +
    aes(y = v2x_libdem) + #BREAK
    aes(size = pop) + #BREAK2
    geom_point(alpha = 0.5) + #BREAK3
    xlab("GDP per capita in 1992, US$") +
    ylab("V-Dem measure of liberal democracy") +
    scale_x_log10(breaks = c(400, 1000,
                             4000, 10000,
                             40000)) +
    labs(title = "Countries richer in 1992 have\nhigher levels of democracy now") +
    labs(caption = "Source: V-Dem, Penn World Tables") + #BREAK4
    scale_colour_discrete("Region") +
    scale_size_continuous(guide = FALSE) + 
    theme_bw() + #BREAK5
    theme(legend.position = "bottom") 
```

---

`r chunk_reveal(chunk_name = "addaes2", break_type = "auto")`

---

# Small multiples


```{r smallmults, eval = FALSE}
ggplot(dat) +
    aes(x = gdp_pc_1992) +
    aes(y = v2x_libdem) + #BREAK
    aes(size = pop) + #BREAK2
    geom_point(alpha = 0.5) + #BREAK3
    xlab("GDP per capita in 1992, US$") +
    ylab("V-Dem measure of liberal democracy") +
    scale_x_log10(breaks = c(400,
                             10000,
                             40000)) +
    labs(title = "Countries richer in 1992 have\nhigher levels of democracy now") +
    labs(caption = "Source: V-Dem, Penn World Tables") + #BREAK4
    scale_colour_discrete("Region") +
    scale_size_continuous(guide = FALSE) +
    facet_wrap(~e_regiongeo) +
    geom_smooth(method = "lm", se = FALSE) + 
    theme_bw() + #BREAK5
    theme(legend.position = "bottom") 
```

---

`r chunk_reveal(chunk_name = "smallmults", break_type = "auto")`

---

# Next steps

 - If you have data which could be presented using a scatter-plot, experiment
 - Pay attention to labelling and captions
 - Think about mapping additional aesthetics. Is it worth it? Does it become too busy?
 
