---
title: "Bar charts"
author: "Chris Hanretty"
date: "November 2020"
output:
  xaringan::moon_reader:
    css: ["default", "rhul.css", "rhul-fonts.css","https://cdnjs.cloudflare.com/ajax/libs/animate.css/3.7.0/animate.min.css"]
    nature:
          highlightLines: true
---


```{r knitropts, echo = FALSE, message = FALSE}

knitr::opts_chunk$set(dpi=300,fig.width=4.5, fig.height = 4, warning = TRUE, message = TRUE, out.width = "100%")

```

# About bar-charts

 - Less common than you might think in the social sciences, but still useful
 - Show how frequent a particular (categorical) outcome is according
   to another (categorical) variable
 - In this case, the categorical outcome is "being a female
   parliamentary candidate", and the categorical variable is
   "political party".
 
---

# About this session

 - This session will show you how to construct a bar-chart using `ggplot`
 - It will also give some guidance on how to **reshape** and **tidy**
   data. You can skip this part if you like.
 - You'll encounter the tidyverse functions `select()`, `mutate()`, `summarize()`
 - You'll also learn about **pivoting** data
 - Useful resources include:
    * Chapter 3 of ["R for Data Science"](https://r4ds.had.co.nz/data-visualisation.html)
	* The [ggplot2 manual](https://ggplot2.tidyverse.org/)
	* Kieran Healy's [Data Visualization](https://socviz.co/)

---

# Load the libraries

```{r loadlibs}
library(tidyverse)
```

First line of code in most of your projects.

```{r loadlibs2, echo = FALSE}
library(flipbookr)
```

---

# Get the data in

The data we're using comes from the British Election Study results
file. 

I'll show you how to plot a bar-chart using a *tidy* version of the data.

Then I'll show you how I tidied the data. 

If you want the tidied data, download from [http://www.foo.org/](http://www.foo.org/)

---

# What the data looks like

```{r datin}
dat <- read.csv("2019_party_gender_count.csv")
head(dat, 10)
```

Note that there are missing values for candidates who haven't yet been coded. 

```{r figcode, echo = FALSE, eval = FALSE, fig = FALSE}
ggplot(dat) +
    aes(x = Party) +
    aes(y = value) +
    aes(fill = CandidateGender) +
    scale_x_discrete("Candidate's party") +
    scale_y_continuous("Number of candidates") +
    geom_col() +
    scale_fill_discrete("Gender") + 
    labs(title = "Male candidates were more common\nthan female candidates in GE2019") +
    labs(caption = "Source: BES results file") +
    theme_bw() +
    theme(legend.position = "bottom")
```

---

`r chunk_reveal(chunk_name = "figcode", break_type = "auto")`

---

# Stacked or dodged bars?

By default, ggplot will **stack** charts which have different aesthetics like colour.

We can change this behaviour by changing the `position` argument to geom_col. 

```{r figcode2, echo = FALSE, eval = FALSE, fig = FALSE}
ggplot(dat) +
    aes(x = Party) +
    aes(y = value) +
    aes(fill = CandidateGender) +
    scale_x_discrete("Candidate's party") +
    scale_y_continuous("Number of candidates") +
    geom_col(position = "dodge") +
    scale_fill_manual("Gender", values = c("darkgrey", "darkred", "grey27", "peachpuff3")) + 
    labs(title = "Male candidates were more common\nthan female candidates in GE2019") +
    labs(caption = "Source: BES results file") +
    theme_bw() +
    theme(legend.position = "bottom")
```

---

`r chunk_reveal(chunk_name = "figcode2", break_type = "auto")`

---

# Tidying BES data

I've shown you how to plot using a **tidy** data-set.

Now I want to show you how to create that tidy data-set by starting
with a real-world dataset which is less tidy.

The data-set I use is the BES results file, available at [https://www.britishelectionstudy.com/wp-content/uploads/2020/02/BES-2019-General-Election-results-file-v1.0.xlsx](https://www.britishelectionstudy.com/wp-content/uploads/2020/02/BES-2019-General-Election-results-file-v1.0.xlsx)

---

# Getting the data in

```{r besin}
library(rio)
bes <- rio::import("https://www.britishelectionstudy.com/wp-content/uploads/2020/02/BES-2019-General-Election-results-file-v1.0.xlsx")
```

The `rio` package is a helpful package for **R** **i**mport and **o**utput of files in different formats.

If you get an error when you try to load `rio`, it's probably because you haven't installed it. Try

```{r eval = FALSE}
install.packages("rio")
```

If you can't install `rio`, you'll need to get someone with admin
privileges to install it on your machine.

---

# Selecting some variables

There are lots of variables in the BES results file. If you don't believe me, try:

```{r ncoleg, echo = TRUE, eval = FALSE}
ncol(bes)
```

If you want to see the (long) list of variable names, try:

```{r nameseg, echo = TRUE, eval = FALSE}
names(bes)
```

Let's select only the variables we're interested in:

```{r selecteg}
bes <- bes %>%
    select(ConstituencyName,
           ConPPCsex19,
           LabPPCsex19,
           LDPPCsex19)
```

---

# Pivoting the data

We have data where each row corresponds to a constituency and where
three columns contain information on a variable (candidate gender)
spread across three categories (parties).

We want to **pivot** this data so that we have data where each row
corresponds to a constituency/party combination, and where a single
variable records information on candidate gender.

This will give us a **longer** data frame, so we use `pivot_longer()`.

---

# Pivot code

```{r pivotcode}
bes <- bes %>%
    pivot_longer(cols = c(ConPPCsex19,
                          LabPPCsex19,
                          LDPPCsex19),
                 names_to = "Party",
                 values_to = "CandidateGender")

head(bes)
```

---

# Summarizing code

We're on the way to our tidy data -- we just need to create counts of
individuals of each gender, by party.

To do this, we use the `group_by` function, and the `n()` function
(which tells us how many observations there are).

```{r summarizecode}
dat <- bes %>%
    group_by(Party, CandidateGender) %>%
    summarize(value = n())

head(dat, 10)
```

---

# Renaming variables

There are two things we might want to do before plotting.

First, we should **sub**stitute the text "PPCsex19" in the variable `Party`.

Second, we could eliminate observations with missing values.

Here's some code that does that

```{r mutatecode}
dat <- dat %>%
    filter(!is.na(CandidateGender)) %>%
    mutate(Party = sub("PPCsex19", "", Party))

head(dat)
```

---

# Next steps

1. Think about how you might plot the same information, but for multiple elections.

2. Think about whether you want to plot raw numbers of proportions, and how you might use `group_by()` and `mutate` to create a new variable with proportions.

3. Think about how to use colour to draw attention to particular categories. Look at sites like [colorbrewer.org](colorbrewer.org) to get contrasting colours which work well with photocopiers and colour-blind viewers.
