---
title: "Visualizations over time"
author: "Chris Hanretty"
date: "November 2020"
output:
  xaringan::moon_reader:
    css: ["default", "rhul.css", "rhul-fonts.css","https://cdnjs.cloudflare.com/ajax/libs/animate.css/3.7.0/animate.min.css"]
    nature:
          highlightLines: true
---


```{r knitropts, echo = FALSE, message = FALSE}

knitr::opts_chunk$set(dpi=300,fig.width=4.5, fig.height = 4, warning = TRUE, message = TRUE, out.width = "100%")

```

# Time and visualization

 - Often, we want to narrate how a quantity has evolved **over time**
 - The classic example of this is support for a particular policy
 - When time is a year, it can be treated as a number;
 - When we have more detailed measures (days and months), it's more complicated
 
---

# About this session

 - This session will show you how to:
    * get R to understand date objects
	* Separate out columns
	* plot using `geom_line()`
 - You'll encounter the tidyverse functions `separate()`
 - You'll learn about `as.Date()` and `as.numeric()`
 - Useful resources include:
    * Chapter 3 of ["R for Data Science"](https://r4ds.had.co.nz/data-visualisation.html)
	* The [ggplot2 manual](https://ggplot2.tidyverse.org/)
	* Kieran Healy's [Data Visualization](https://socviz.co/)

---

# Load the libraries

```{r loadlibs}
library(tidyverse)
```

First line of code in most of your projects.

```{r loadlibs2, echo = FALSE}
library(flipbookr)
```

---

# Get the data in

The data we're using comes from the Wikipedia page, [Opinion polling on Scottish independence](https://en.wikipedia.org/wiki/Opinion_polling_on_Scottish_independence#Post-referendum_polling)

I've got a CSV file where I've removed some rows which refer to
specific events, but otherwise the data is as you would see it on the
website.

If you want the tidied data, download from [https://raw.githubusercontent.com/dataviz/main/02_candidacy/2019_party_gender_count.csv](https://raw.githubusercontent.com/dataviz/main/02_candidacy/2019_party_gender_count.csv)

---

# What the data looks like

```{r datin}
dat <- read.csv("indy_polling.csv")
head(dat, 10)
```

---

# Problems with the data

1. The fieldwork gives a range of dates, not individual dates
2. The values in `Yes` and `No` are percentages, not numbers

We can fix these problems!

---

# Separating columns

```{r cleancode}
dat <- dat %>%
    separate(col = Fieldwork,
             into = c("FwkStart", "FwkStop"),
             sep = "–")

head(dat)
```

Note that I had to copy and paste the "hypen" (annoyingly it's an em-dash, not a hyphen).

---

# Converting things to dates

```{r cleandates}
dat <- dat %>%
    mutate(FwkStart = as.Date(FwkStart,
                             format = "%d %b %Y"),
           FwkStop = as.Date(FwkStop,
                             format = "%d %b %Y")) %>%
    mutate(Date = if_else(is.na(FwkStop),
                         FwkStart,
                         FwkStop))

head(dat)
```

---

# Formatting codes

**%a**
Abbreviated weekday name in the current locale on this platform.

**%A**
Full weekday name in the current locale. 

**%b**
Abbreviated month name in the current locale on this platform.

**%B**
Full month name in the current locale. 

**%d**
Day of the month as decimal number (01--31).

**%m**
Month as decimal number (01--12).

**%y**
Year without century (00--99). 

**%Y**
Year with century.


See full list at [https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/strptime](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/strptime) or Google `date formatting R`.

---

# Tidying up numbers

We'll use the `sub()` function which we've used before, and the
`as.numeric()` function, which we haven't.

```{r asnumericcode}
dat <- dat %>%
    mutate(Yes = gsub("%", "", Yes),
           Yes = as.numeric(Yes),
           No = gsub("%", "", No),
           No = as.numeric(No))
```

```{r plottingcode, echo = FALSE, eval = FALSE}
ggplot(dat) +
    aes(x = Date) +
    aes(y = Yes) +
    scale_x_date() +
    scale_y_continuous(limits = c(0, 60)) +
    xlab("Date") +
    ylab("Support for independence") +
    geom_point() +
    geom_smooth() +
    geom_vline(xintercept = as.Date("2016-06-23",
                                    format = "%Y-%m-%d"),
               colour = "darkgrey") + 
    labs(title = "Support for independence since 2014") +
    labs(subtitle = "Vertical line shows Brexit referendum;\nblue line shows trend") + 
    labs(caption = "Source: Wikipedia") +
    theme_bw()
```

---

`r chunk_reveal(chunk_name = "plottingcode", break_type = "auto")`

---

# Next steps

1. How could you `pivot` the data so that the data has one row for each combination of polling and Yes/No
2. Are there other aesthetics you might want to map to the plotted points?
